---
title : 'Basic Concepts of Mixture Models'
subtitle : 'Bayesian Statistics: Mixture Models'
categories:
  - Bayesian Statistics
keywords:
  - Mixture Models
  - Homework Honors
---

## HW+ - Sim mixture of exponential distributions


::: {.callout-note}
### Instructions

- Modify code to Generate n observations from a mixture of two Gaussian # distributions into code to sample 100 random numbers from a mixture of 4 exponential distributions with means 1, 4, 7 and 10 and weights 0.3, 0.25, 0.25 and 0.2, respectively. 
- Use these sample to approximate the mean and variance of the mixture.

:::

```{r}
#| label: C3-L01-Ex6-1
set.seed(238)                            # reproducibility
n <- 100                                  # the sample size
w <- c(0.3, 0.25, 0.25,  0.2)        # weights
lambda <- c(1, 4, 7 , 10)                 # means
rates <- 1 / lambda                       # inverses of the means
ac <- sample (1:length(w), n, replace=T, prob =w) # smaple the active component
x <- rexp(n,rate = rates[ac])            # sample from the exponential distribution
```

next we use the samples to estimate the mean and variance of the mixture

```{r}
#| label: C3-L01-Ex6-2
(sample_mean=mean(x))
(sample_variance=var(x))
```
