---
title : 'Estimating the number of components in Bayesian settings'
subtitle : 'Bayesian Statistics: Mixture Models'
categories:
  - Bayesian Statistics
keywords:
  - Mixture Models
  - Homework
  - Number of Components
  - Bayesian Estimation
---

## HW - Estimating the number of components in Bayesian settings

1. Let $K^∗$ be the prior expected number of occupied components in a mixture model with $K$ components where the weights are given a Dirichlet prior $(w_1, \ldots, w_K) \sim \text{Dir}(2K, \ldots, 2K)$. If you have $n = 400$ observations, what is the expected number of occupied components, $E(K^∗)$ according to the exact formula we discussed in the lecture? Round your answer to one decimal place.

```{r}
n      = 400
alpha  = 2

EKstar = 0
for(i in 1:n){
  EKstar = EKstar + alpha/(alpha + i -1)
}
print(EKstar)
```
11.1

2. Consider the same setup as the previous question, what is the expected number of occupied components, $E(K^∗)$ according to the exact formula we discussed in the lecture if $n = 100$ instead? Round your answer to one decimal place.

```{r}
n      = 100
alpha  = 2

EKstar = 0
for(i in 1:n){
  EKstar = EKstar + alpha/(alpha + i -1)
}
print(EKstar)
```

8.4

3. What would be the answer to the previous question if you used the approximate formula instead of the exact formula? Remember to round your answer to one decimal place.

```{r}
n      = 100
alpha  = 2
alpha*log((n+alpha-1)/alpha)
```

7.8

4. If you have $n = 200$ observations and a priori expect the mixture will have about 2 occupied components (i.e., $E(K^∗) \approx 2$ a priori), what value of $\alpha$ should you use for the prior $(w_1, \ldots, w_K) \sim \text{Dir}(\alpha K, \ldots, \alpha K)$. Use the approximation $E(K^∗) \approx \alpha \log\left(\frac{n + \alpha - 1}{\alpha}\right)$ to provide an answer, which should be rounded to two decimal

```{r}
n      = 200
alpha  = 2
alpha*log((n+alpha-1)/alpha)
```

0.31