---
title : 'M5HW2 - Estimating the number of components in Bayesian settings'
subtitle : 'Bayesian Statistics: Mixture Models'
categories:
  - Bayesian Statistics
keywords:
  - Mixture Models
  - Homework
  - Number of Components
  - Bayesian Estimation
---

::: {#exr-1}
Let $K^∗$ be the prior expected number of occupied components in a mixture model with $K$ components where the weights are given a Dirichlet prior $(w_1, \ldots, w_K) \sim \text{Dir}(2K, \ldots, 2K)$. If you have $n = 400$ observations, what is the expected number of occupied components, $E(K^∗)$ according to the exact formula we discussed in the lecture? Round your answer to one decimal place.
:::

::: {.callout-note .solution collapse="true"}
## Solution

To compute the expected number of occupied components, we can use the formula:

$$
E(K^*) = \sum_{i=0}^n \frac{\alpha}{\alpha+i-1}
$$
Thus, the expected number of occupied components is given by:


```{r}
#| label: exr-1-sol

E.Kstar = 0
for (i in 1:400){
  E.Kstar = E.Kstar +  2/(2+i-1)
}
round(E.Kstar,1)

```

11.1
:::


::: {#exr-2}
Consider the same setup as the previous question, what is the expected number of occupied components, $E(K^∗)$ according to the exact formula we discussed in the lecture if $n = 100$ instead? Round your answer to one decimal place.
:::

::: {.callout-note .solution collapse="true"}
## Solution
```{r}
n      = 100
alpha  = 2

EKstar = 0
for(i in 1:n){
  EKstar = EKstar + alpha/(alpha + i -1)
}
print(EKstar)
```

8.4
:::

::: {.callout-note .solution collapse="true"}
## Solution
What would be the answer to the previous question if you used the approximate formula instead of the exact formula? Remember to round your answer to one decimal place.

:::

::: {#exr-3-sol}
```{r}
#| label: exr-3-sol

alpha = 2
n = 100
Kstar = alpha*log((n+alpha-1)/alpha)
round( Kstar,1)
```

7.8
:::

::: {#exr-4}
If you have $n = 200$ observations and a priori expect the mixture will have about 2 occupied components (i.e., $E(K^∗) \approx 2$ a priori), what value of $\alpha$ should you use for the prior $(w_1, \ldots, w_K) \sim \text{Dir}(\alpha K, \ldots, \alpha K)$. Use the approximation $E(K^∗) \approx \alpha \log\left(\frac{n + \alpha - 1}{\alpha}\right)$ to provide an answer, which should be rounded to two decimal
:::

::: {.callout-note .solution collapse="true"}
## Solution

```{r}
n      = 200
alpha  = 2
alpha*log((n+alpha-1)/alpha)
```

0.31
:::