---
date: 2024-11-09
title: "Bayesian Inference in the NDLM: Part 2"
subtitle: Time Series Analysis
description: "Normal Dynamic Linear Models (NDLMs) are a class of models used for time series analysis that allow for flexible modeling of temporal dependencies."
categories: 
  - Bayesian Statistics
  - Time Series
keywords: 
  - Time series
  - Filtering
  - Smoothing
  - NDLM
  - Normal Dynamic Linear Models
  - Seasonal NDLM
  - Superposition Principle
  - R code
fig-caption: Notes about ... Bayesian Statistics
title-block-banner: images/banner_deep.jpg
---

## Filtering, Smoothing and Forecasting: Unknown observational variance  :movie_camera: {#section-NDLM-unknown-variance}

::: {.callout-caution collapse="true"}
## Bumpy ride ahead! {.unnumbered}

- This material seems unnecessarily complicated:

1. This section seems like a lots of mathematical gobbledygook without any context of motivation for application for each of the equations being presented.
  - After writing everything out and reviewing it multiple times, I began to develop a purely mathematical intuition i.e. I can relate them to the the general Bayesian inference framework,
  - I can't however say I am much closer to using them in engineering solution based on this mathematics.
  - I doubt that doing a single capstone project will make much of a difference.

2. While we will see three application in this module, (Nile river data, google trends, and EEG data). However, what I need to understand for each of these filtering, smoothing and forecasting distributions why are they useful rather than the original ones. I mean when we use to solve a specific challenge, for a dataset. 

3. Perhaps the root cause here is that this course fails to acknowledge the connection of NDLM to the Kalman Filters and the much wider literature on Kalman Filters than NDLM. It appears to me that in the first course of the Kalman filtering specialization (KF bootcamp) there are a number of application that are presented each with a different requirements and that if we map the different challenges to the NDLM framework we should be able to navigate the math more easily.

4. Anyhow a lookup indicates that this model is called in the state-space modeling terminology as **local level model** for local level or trend filtering, such as in structural time series models, and in the  Kalman filtering  terminology as a KF with variance learning.
:::

In the case of the NDLM, where we **assume** That the observational variance $\nu_t=v$ i.e. is constant over time.

### Inference in the NDLM with unknown constant observational variance:

Let $\nu_t = v\ \forall t$ with $\nu$ unknown and consider a DLM with the following structure:
$$
\begin{aligned}
  y_t &= \mathbf{F}_t^\top \boldsymbol{\theta}_t + \nu_t,                
      &\nu_t &\sim \mathcal{N} (0, \nu) 
      &\text{(obs)}\\
  \boldsymbol{\theta}_t &= \mathbf{G}_t \boldsymbol{\theta}_{t-1} + \boldsymbol{\omega}_t, 
      & \boldsymbol{\omega}_t 
      &\sim \mathcal{N} (0, \nu \mathbf{W}^*_t) 
      &\text{(sys)}
\end{aligned}
$$ {#eq-NDLM-unknown-variance}

- where we have the usual suspects:
  - An observation equation, relating the observed $y_t$ to the hidden state $\theta_t$ plus some sensor noise $\nu_t$.
  - A state equation, which describes how the hidden state $\theta_t$ evolves plus some process noise $\boldsymbol{\omega}_t$, 
  - $\mathbf{F}_t$  is a *known* observation operator,
  -  $\mathbf{G}_t$ is a *known* state transition operator,
  - $\boldsymbol{\theta}_t$ is the hidden state vector, and
  - $\boldsymbol{\omega}_t$ is the process noise, e.g. friction or wind.
  - $\mathbf{W}_t$ is a *known* covariance matrix, for process noise.


::: {.callout-important collapse="False"}
## Question: why does $\nu$ appear in the $w_t$ distribution equation? {.unnumbered}

Prado says is also **assuming** that the system noise is conditional on the variance $\nu$.
Yet this seems to be an assumption that should be justified.

In fact, What we are missing is a logical Model for this model i.e. simple use case

:::

with conjugate prior distributions:
$$
\begin{aligned}
(\boldsymbol{\theta}_0 \mid \mathcal{D}_0, \nu) &\sim \mathcal{N} (\mathbf{m}_0, \nu\ \mathbf{C}^*_0) \\ (\nu \mid \mathcal{D}_0) &\sim \mathcal{IG}\left(\frac{n_0}{2}, \frac{d_0}{2}\right ) \\
 d_0 &= n_0 s_0
\end{aligned}
$$ {#eq-NDLM-unknown-variance-priors}

### Filtering

We now consider filtering updating uncertainty. Which lets us infer current time step parameters or observation based on the previous time step's information and $\nu$ as being normal distributed.

we will be interesting the following distributions:

- $p(\boldsymbol{\theta}_t \mid \mathcal{D}_{t−1})$ the prior density for the state vector at time t given information up to the preceding time;
- $p(y_t \mid \mathcal{D}_{t−1})$ the one-step-ahead predictive density for the next observation;
- $p(\boldsymbol{\theta}_t \mid \mathcal{D}_t)$ the posterior density for the state vector at time t given $\mathcal{D}_{t−1}$ and $y_t$;
- the h-step-ahead forecasts $p(y_{t+h} \mid \mathcal{D}_t)$ and $p(\boldsymbol{\theta}_{t+h} \mid \mathcal{D}_t)$;
- $p(\boldsymbol{\theta}_t \mid \mathcal{D}_T)$ the smoothing density for $\boldsymbol{\theta}_t$ where $T > t$.

Assuming:

$$
(\boldsymbol{\theta}_{t-1} \mid \mathcal{D}_{t-1}, \nu) \sim \mathcal{N} (\mathbf{m}_{t-1}, \nu\ \mathbf{C}^*_{t-1})
$$ 

By filtering we obtain the next time step parameters conditional on the previous time step's information and the variance $\nu$ as being normal distributed.

$$
\begin{aligned}
(\boldsymbol{\theta}_t \mid \mathcal{D}_{t-1}, \nu) &\sim \mathcal{N} (\mathbf{a}_t, \nu \mathbf{R}^*_t) \\
  \mathbf{a}_t &= \mathbf{G}_t \mathbf{m}_{t-1} \\
  \mathbf{R}^*_t &= \mathbf{G}_t \mathbf{C}^*_{t-1} \mathbf{G}'_t + \mathbf{W}^*_t
\end{aligned}
$$ {#eq-NDLM-unknown-variance-filtering}


We can also marginalize $\nu$ by integrating it out, we have parameters that are unconditional on $\nu$:

$$
\begin{aligned}
  (\boldsymbol{\theta}_t \mid \mathcal{D}_{t-1}) &\sim T_{n_{t-1}} (\mathbf{a}_t, \mathbf{R}_t) \\
  \mathbf{R}_t &\equiv s_{t-1} \mathbf{R}^*_t
\end{aligned}
$$ {#eq-NDLM-unknown-variance-filtering-unconditional}



with $s_t\ \forall t$ given in @eq-NDLM-unknown-variance-filtering.

- $(y_t \mid \mathcal{D}_{t-1}, \nu) \sim \mathcal{N} (f_t, \nu q^*_t)$
  - with 
    - $f_t = F'_t \mathbf{a}_t$ 
    - $q^*_t = (1 + F'_t \mathbf{R}^*_t F_t)$ 
    
and unconditional on $\nu$ we have 

- $(y_t \mid \mathcal{D}_{t-1}) \sim T_{n_{t-1}} (f_t, q_t)$
  - with 
    - $q_t = s_{t-1} q^*_t$

- $(v \mid \mathcal{D}_t) \sim \mathcal{IG}\left( \tfrac{n_t}{2}, \tfrac{s_t}{2}\right )$
  -  with
    - $n_t = n_{t-1} + 1$
  $$ 
  s_t = s_{t-1} + \frac{s_{t-1}}{n_t} \left ( \frac{e^2_t}{q^*_t} - 1 \right ), 
  $$ {#eq-NDLM-unknown-variance-filtering}

  here $e_t = y_t - f_t$

- $(\theta_t \mid D_t, v) \sim N (m_t, vC^*_t)$, 
  - with
    - $m_t = a_t + A_t e_t$, and 
    - $C^*_t = R^*_t - A_t A'_t q^*_t$.

Similarly, unconditional on $\nu$ we have

- $(\boldsymbol{\theta}_t\mid \mathcal{D}_t) \sim T_{n_t} (\mathbf{m}_t, \mathbf{C}_t)$
  - with 
    - $\mathbf{C}_t = s_t \mathbf{C}^*_t$

### Forecasting

Similarly, we have the forecasting distributions:

- $(\boldsymbol{\theta}_{t+h} \mid \mathcal{D}_t) \sim T_{n_t} (\mathbf{a}_{t}(h), \mathbf{R}_{t}(h))$,
- $(y_{t+h} \mid \mathcal{D}_t) \sim T_{n_t} (f_{t}(h), q_{t}(h))$,

- with
  - $\mathbf{a}_{t}(h) = \mathbf{G}_{t+h} \mathbf{a}_{t}(h-1)$,
  - $\mathbf{a}_{t}(0) = \mathbf{m}_t$, and

$$
\begin{aligned}
  \mathbf{R}_{t}(h) &= \mathbf{G}_{t+h} \mathbf{R}_{t}(h-1) \mathbf{G}'_{t+h} + \mathbf{W}_{t+h} \\
  \mathbf{R}_{t}(0) &= \mathbf{C}_t
\end{aligned}
$$

$$
f_{t}(h) = \mathbf{F}'_{t+h} \mathbf{a}_{t}(h) \qquad q_{t}(h) = \mathbf{F}'_{t+h} \mathbf{R}_{t}(h) \mathbf{F}_{t+h} + s_{t}
$$


### Smoothing


Finally, the smoothing distributions have the form:

$$
(\boldsymbol{\theta}_t \mid \mathcal{D}_T) \sim T_{n_T} (a_T (t - T), R_T (t - T) s_T / s_t)
$$

- where
  -$a_T (t - T)$ and $R_T (t - T)$ 
with
$$
a_T (t - T) = m_T - B_T [a_{T+1} - a_T (t - T + 1)]
$$

$$
R_T (t - T) = C_T - B_T [R_{T+1} - R_T (t - T + 1)] B'_T
$$

with 

$$
B_T = C_T G'_{T+1} R^{-1}_{T+1}, \quad a_T (0) = m_T , \quad R_T (0) = C_T.
$$

::: {.callout-note collapse="true"}
## Video Transcript {.unnumbered}

{{< include transcripts/c4/04_week-4-normal-dynamic-linear-models-part-ii/02_bayesian-inference-in-the-ndlm-part-ii/01_filtering-smoothing-and-forecasting-unknown-observational-variance.en.txt >}}
:::

## Summary of Filtering, Smoothing and Forecasting Distributions, NDLM unknown observational variance  :spiral_notepad: 



## Specifying the system covariance matrix via discount factors  :movie_camera: 

::: {.callout-note collapse="true"}
## Video Transcript {.unnumbered}

{{< include transcripts/c4/04_week-4-normal-dynamic-linear-models-part-ii/02_bayesian-inference-in-the-ndlm-part-ii/03_specifying-the-system-covariance-matrix-via-discount-factors.en.txt >}}

:::

## NDLM, Unknown Observational Variance: Example  :movie_camera: 

This is a walk though of the R code in @lst-NDLM-unknown-variance

::: {.callout-note collapse="true"}
## Video Transcript {.unnumbered}

{{< include transcripts/c4/04_week-4-normal-dynamic-linear-models-part-ii/02_bayesian-inference-in-the-ndlm-part-ii/04_ndlm-unknown-observational-variance-example.en.txt >}}

:::

## code: NDLM, Unknown Observational Variance Example  :spiral_notepad: $\mathcal{R}$

This code allows time-varying $F_t$, $G_t$ and $W_t$ matrices and assumes an unknown but constant $\nu$. 
It also allows the user to specify $W_t$ using a discount factor $\delta \in (0,1]$ or assume $W_t$  known.


### Nile River Level Filtering Smoothing and Forecasting Example

```{r}

source('all_dlm_functions_unknown_v.R')
source('discountfactor_selection_functions.R')

## Example: Nile River Level (in 10^8 m^3), 1871-1970 
## Model: First order polynomial DLM
plot(Nile) 
n=length(Nile) #n=100 observations 
k=5
T=n-k
data_T=Nile[1:T]
test_data=Nile[(T+1):n]
data=list(yt = data_T)


## set up matrices for first order polynomial model 
Ft=array(1, dim = c(1, 1, n))
Gt=array(1, dim = c(1, 1, n))
Wt_star=array(1, dim = c(1, 1, n))
m0=as.matrix(800)
C0_star=as.matrix(10)
n0=1
S0=10

## wrap up all matrices and initial values
matrices = set_up_dlm_matrices_unknown_v(Ft, Gt, Wt_star)
initial_states = set_up_initial_states_unknown_v(m0, 
                                      C0_star, n0, S0)

## filtering 
results_filtered = forward_filter_unknown_v(data, matrices, 
                                            initial_states)
ci_filtered=get_credible_interval_unknown_v(results_filtered$mt, 
                                    results_filtered$Ct, 
                                     results_filtered$nt)

## smoothing
results_smoothed=backward_smoothing_unknown_v(data, matrices, 
                                             results_filtered)
ci_smoothed=get_credible_interval_unknown_v(results_smoothed$mnt, 
                                         results_smoothed$Cnt, 
                                         results_filtered$nt[T])

## one-step ahead forecasting
results_forecast=forecast_function_unknown_v(results_filtered, 
                                                k,  matrices)
ci_forecast=get_credible_interval_unknown_v(results_forecast$ft, 
                                          results_forecast$Qt, 
                                     results_filtered$nt[T])
```

```{r}
#| label: fig-NDLM-unknown-variance-nile-data
#| fig-cap: Nile River Level Filtering, Smoothing and Forecasting Results


## plot results
index=seq(1871, 1970, length.out = length(Nile))
index_filt=index[1:T]
index_forecast=index[(T+1):(T+k)]

plot(index, Nile, main = "Nile River Level ",type='l',
     xlab="time",ylab="feet",lty=3,ylim=c(400,1500))
points(index,Nile,pch=20)

lines(index_filt,results_filtered$mt, type='l', col='red',lwd=2)
lines(index_filt,ci_filtered[, 1], type='l', col='red', lty=2)
lines(index_filt,ci_filtered[, 2], type='l', col='red', lty=2)
lines(index_filt,results_smoothed$mnt, type='l', col='blue',lwd=2)
lines(index_filt, ci_smoothed[, 1], type='l', col='blue', lty=2)
lines(index_filt, ci_smoothed[, 2], type='l', col='blue', lty=2)

lines(index_forecast, results_forecast$ft, type='l', 
      col='green',lwd=2)
lines(index_forecast, ci_forecast[, 1], type='l', 
      col='green', lty=2)
lines(index_forecast, ci_forecast[, 2], type='l', 
      col='green', lty=2)
legend("topright",
       legend = c("Observed", "Filtered mean", "Filtered CI", 
                  "Smoothed mean", "Smoothed CI", 
                  "Forecast mean", "Forecast CI"),
       col = c("black", "red", "red", "blue", "blue", "green", "green"),
       lty = c(NA, 1, 2, 1, 2, 1, 2),
       pch = c(20, NA, NA, NA, NA, NA, NA),
       lwd = c(NA, 2, 1, 2, 1, 2, 1),
       bty = "n",
       ncol = 2)



```

::: {.callout-note collapse="false"}
## Overthinking the Nile River Data {.unnumbered}

We use the Nile dataset throughout the course, despite the Nile data being notorious for having long term dependencies and being highly non-linear, which might present many challenges for the NDLM framework. c.f. the joint work of two giants [Benoit Mandelbrot](https://en.wikipedia.org/wiki/Benoit_Mandelbrot) and [Harold Edwin Hurst](https://en.wikipedia.org/wiki/Harold_Edwin_Hurst).
Which led to the development of the Hurst exponent and the [fractional Brownian motion](https://en.wikipedia.org/wiki/Fractional_Brownian_motion), i.e. fractional Gaussian noise and fractional ARIMA [ARFIMA](https://en.wikipedia.org/wiki/Autoregressive_fractionally_integrated_moving_average) which are more suitable for modeling the Nile data but not covered in this course.

Just a couple of concepts here:

In Fractional Brownian Motion (fBm) draws from the Normal need not be independent. With the concept of Hurst exponent $H$ which is a measure of the long-term memory of a time series. The Hurst exponent is defined as follows:
$$
\mathbb{E}[B_{H}(t)B_{H}(s)]={\tfrac {1}{2}}(|t|^{2H}+|s|^{2H}-|t-s|^{2H}),
$$ {#eq-fbm-autocovariance}

- where:
  - $B_{H}(t)$ is the fractional Brownian motion, and
  - $H$ is the Hurst exponent.

2. in ARFIMA, the autocovariance we allow the exponent of the lag $k$ to be a real number $d$ rather than an integer, i.e. we allow for fractional differencing via a formal binomial series expansion of the form:

$$\begin{array}{c} {\displaystyle {\begin{aligned}(1-B)^{d}&=\sum _{k=0}^{\infty }\;{d \choose k}\;(-B)^{k}\\&=\sum _{k=0}^{\infty }\;{\frac {\prod _{a=0}^{k-1}(d-a)\ (-B)^{k}}{k!}}\\&=1-dB+{\frac {d(d-1)}{2!}}B^{2}-\cdots \,.\end{aligned}}} \end{array}
$$ {#eq-ARFIMA}

:::

## Case Study EEG data  :movie_camera:  {#section-NDLM-eeg}


```{r}
### EEG example ##################
source('all_dlm_functions_unknown_v.R')
source('discountfactor_selection_functions.R')

eeg_data <- scan("data/eeg.txt")
length(eeg_data)
eeg_data_gh <- scan("data/eeg_gh.txt")
length(eeg_data_gh)
plot(eeg_data, type='l', main="EEG Data", xlab="Time", ylab="EEG Signal")
head(eeg_data)
head(eeg_data_gh)

T <- length(eeg_data)
data <- list(yt = eeg_data[])
```

::: {.callout-note collapse="true"}
## Video Transcript

{{< include transcripts/c4/04_week-4-normal-dynamic-linear-models-part-ii/03_case-studies/01_eeg-data.en.txt >}}

:::

## Case Study: Google Trends :movie_camera:  {#section-NDLM-google-trends}

```{r}
```



::: {.callout-note collapse="true"}
## Video Transcript

{{< include transcripts/c4/04_week-4-normal-dynamic-linear-models-part-ii/03_case-studies/02_google-trends.en.txt >}}

:::