---
title: "Probability - M1L1"
subtitle: "Capstone Project: Bayesian Conjugate Analysis for Autogressive Time Series Models"
subject: "Bayesian Statistics"
description: "Outline of probability"
categories:
  - Bayesian Statistics
  - Capstone Project
keywords:
  - Odds
  - Probability
  - Expectation
  - σ-Algebra
  - Probability Measure
  - Probability Space
  - Probability Paradigms
  - Coherence
  - Abraham de Moivre
  - Bruno de Finetti
  - Objectivity
  - Variance
  - Covariance
  - Correlation
---

::: {.callout-note collapse="true"}
## Learning Objectives

-   [x] Explain the procedure for sampling from the posterior distribution, and code in R to achieve it.
-   [x] Calculate the posterior distributions of model parameters.
-   [x] Describe the AR models under Bayesian conjugate setting. Specify the likelihood function and prior distribution.
:::

## Introduction 🎥 {#sec-capstone-introduction} 

This is a capstone project for the Bayesian Statistics course. 
The instructor is a doctoral student Chi Joe Kao. 
We will do a capstone project.
We will also cover a new type of model - a mixture version of the autoregressive model from the previous course.
We will cover the Bayesian conjugate analysis for autoregressive time series models.

## Prerequisite skill checklist 📖


:::: {.callout-note collapse="true"}
## Prerequisite skill checklist

### Bayesian Statistics
- [x] Interpret and specify the components of Bayesian statistical models: likelihood, prior, posterior.
- [x] Explain the basics of sampling algorithms, including sampling from
standard distributions and using MCMC to sample from non-standard
posterior distributions.
- [x] Assess the performance of a statistical model and compare competing
models using posterior samples.
- [x] Coding in R to achieve the above tasks.

### Mixture Models
- [x] Define mixture model.
- [x] Explain the likelihood function associated with a random sample
from a mixture distribution.
- [x] Derive Markov chain Monte Carlo algorithms for fitting mixture models.
- [x] Coding in R to achieve the above tasks.

### Time Series Analysis
- [x] Define time series and stochastic processes (univariate, discrete-time, equally-spaced)
- [x] Define strong and weak stationarity
- [x] Define the auto-covariance function, the auto-correlation function
(ACF), and the partial autocorrelation function (PACF)
- [x] Definition of the general class of autoregressive processes of order p.
:::

## Read Data 📖 ℛ

### Earth Quake Data


```{r}

## read data, you need to make sure the data file is in your current working directory 
earthquakes.dat <- read.delim("data/earthquakes.txt")
earthquakes.dat$Quakes=as.numeric(earthquakes.dat$Quakes)


y.dat=earthquakes.dat$Quakes[1:100] ## this is the training data
y.new=earthquakes.dat$Quakes[101:103] ## this is the test data
```

### Google Search Index Data


```{r}
## read data, you need to make sure the data file is in your current working directory 
covid.dat <- read.delim("data/GoogleSearchIndex.txt")
covid.dat$Week=as.Date(as.character(covid.dat$Week),format = "%Y-%m-%d")




y.dat=covid.dat$covid[1:57] ## this is the training data
y.new=covid.dat$covid[58:60] ## this is the test data
```
