---
title: "Southern Oscillation Index"
date: "2024-09-01"
keywords: ["enso", "SOI"]
---

## SOI



https://www.ncei.noaa.gov/access/monitoring/enso/soi/


> The Southern Oscillation Index (SOI) is a standardized index based on the observed sea level pressure (SLP) differences between Tahiti and Darwin, Australia. The SOI is one measure of the large-scale fluctuations in air pressure occurring between the western and eastern tropical Pacific (i.e., the state of the Southern Oscillation) during El Niño and La Niña episodes. In general, smoothed time series of the SOI correspond very well with changes in ocean temperatures across the eastern tropical Pacific. The negative phase of the SOI represents below-normal air pressure at Tahiti and above-normal air pressure at Darwin.

let's load the soi data and plot it.


```{r}
#| label: dependencies
if(!require('ocedata')) {
  install.packages('ocedata')
  library('ocedata')
}
```

```{r}
data(soi, package="ocedata")
recent <- subset(soi, year > 1950)
plot(recent$year, recent$index, type='l', xlab="Year", ylab="SOI")
```