these are some datasets used in the specialization as well as some extras


| File Name | Description | Source |
| --- | --- | --- |
| CO2.csv | Mauna Loa CO2 data set | https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/co2.html
| nile.csv | The Nile River data set | https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/Nile
| MEDDAYONMARUS.csv | Housing Inventory: Median Days on Market in the United States (MEDDAYONMARUS)| https://fred.stlouisfed.org/series/MEDDAYONMARUS|
| POILWTIUSDM.csv | Global price of WTI Crude (POILWTIUSDM) | https://fred.stlouisfed.org/series/POILWTIUSDM


## co_2 data set {#sec-co2-data}

```{r}
#| label: lst-just-load-the-co2-data
require(graphics)
plot(co2, ylab = expression("Atmospheric concentration of CO"[2]),
     las = 1)
title(main = "co2 data set")
```

## Nile River data set {#sec-nile-data}
```{r}
#| label: lst-just-load-the-nile-data
require(graphics)
plot(Nile, ylab = "Flow (billion cubic meters)",
     las = 1)
title(main = "Nile River data set")
```


```{r}
#| label: lst-acf-pacf-nile-data
require(stats)
require(graphics)
par(mfrow = c(2, 2))
plot(Nile)
acf(Nile)
pacf(Nile)
ar(Nile) # selects order 2
cpgram(ar(Nile)$resid)
par(mfrow = c(1, 1))
arima(Nile, c(2, 0, 0))

## Now consider missing values, following Durbin & Koopman
NileNA <- Nile
NileNA[c(21:40, 61:80)] <- NA
arima(NileNA, c(2, 0, 0))
plot(NileNA)
pred <-
   predict(arima(window(NileNA, 1871, 1890), c(2, 0, 0)), n.ahead = 20)
lines(pred$pred, lty = 3, col = "red")
lines(pred$pred + 2*pred$se, lty = 2, col = "blue")
lines(pred$pred - 2*pred$se, lty = 2, col = "blue")
pred <-
   predict(arima(window(NileNA, 1871, 1930), c(2, 0, 0)), n.ahead = 20)
lines(pred$pred, lty = 3, col = "red")
lines(pred$pred + 2*pred$se, lty = 2, col = "blue")
lines(pred$pred - 2*pred$se, lty = 2, col = "blue")

## Structural time series models
par(mfrow = c(3, 1))
plot(Nile)
## local level model
(fit <- StructTS(Nile, type = "level"))
lines(fitted(fit), lty = 2)              # contemporaneous smoothing
lines(tsSmooth(fit), lty = 2, col = 4)   # fixed-interval smoothing
plot(residuals(fit)); abline(h = 0, lty = 3)
## local trend model
(fit2 <- StructTS(Nile, type = "trend")) ## constant trend fitted
pred <- predict(fit, n.ahead = 30)
## with 50% confidence interval
ts.plot(Nile, pred$pred,
        pred$pred + 0.67*pred$se, pred$pred -0.67*pred$se)

## Now consider missing values
plot(NileNA)
(fit3 <- StructTS(NileNA, type = "level"))
lines(fitted(fit3), lty = 2)
lines(tsSmooth(fit3), lty = 3)
```


### ARFIMA using `fracdiff` package

```{r}
#| label: lst-nile-data-arfima-fracdiff
library(longmemo)
# Load the Nile data:
library(graphics)
nile <- Nile
## fit a model to the Nile data
#fit <- auto.arima(nile, d=c(0,1), approximation=FALSE, stepwise=FALSE)
#summary(fit)

library(fracdiff)
fit <- fracdiff(nile, nar=1, nma=1)
summary(fit)

# View the model summary
summary(fit)
# Plot the residuals
plot(residuals(fit))
# Check the ACF and PACF of residuals
acf(residuals(fit))
pacf(residuals(fit))

library(forecast)
forecast_nile <- forecast(fit, h=20)
plot(forecast_nile)


Box.test(residuals(fit), lag=20, type="Ljung-Box")
```

### ARFIMA using `forecast` package


```{r}
#| label: lst-nile-data-arfima-forecast
library(forecast)

# Fit an ARFIMA(p, d, q) model to the Nile data
fit <- arfima(Nile)

# View summary
summary(fit)

# Access coefficients
fit$coef

# Plot residuals
plot(residuals(fit))

# ACF and PACF of residuals
acf(residuals(fit))
pacf(residuals(fit))

# Forecast
forecast_nile <- forecast(fit, h=20)
plot(forecast_nile)

# Ljung-Box test
Box.test(residuals(fit), lag=20, type="Ljung-Box")

```

