
## Sample code for EM example 1  :spiral_notepad:  $\mathcal{R}$ {#sec-em-code-example-1}

We start in @lst-em-loc-mixture-setup with some house cleaning by clearing the environment and setting the seed for reproducibility.

```{r}
#| label: lst-em-loc-mixture-setup
#| lst-label: lst-em-loc-mixture-setup
#| lst-cap: "Setup"

rm(list=ls())    # <1>
set.seed(81196)  # <2> 
```

1. Clear the environment and load required libraries
2. Set the seed for reproducibility of the results.


Since the algorithm will be tested on simulated data, we proceed in @lst-em-example-synthesize-data to generate the synthetic dataset from a mixture of 2 Gaussians. 

```{r}
#| label: lst-em-example-synthesize-data
#| lst-label: lst-em-example-synthesize-data
#| lst-cap: "Generate data from a mixture with 2 components"

## Ground Truth parameters initialization
KK         = 2          # <1>
w.true     = 0.6        # <2>
mu.true    = rep(0, KK) # <3>
mu.true[1] = 0          # <4>
mu.true[2] = 5          # <4>
sigma.true = 1          # <5>

n  = 120         # <6>
cc = sample(1:KK, n, replace=T, prob=c(w.true,1-w.true)) # <7>
x  = rep(0, n)   # <8>

for(i in 1:n){ 
  # sample from a distribution with mean selected by component indicator
  # the SD is the same for all components as this is a location mixture
  x[i] = rnorm(1, mu.true[cc[i]], sigma.true)
}
```
1. Number of components of the mixture
2. Ground Truth (GT) True weights associated with the components
3. Initialize the true means list
4. Ground Truth (GT) True means for each component
5. GT standard deviation of all components
6. Number of synthetic samples to generate
7. Simulate the latent variables for the component indicator function
8. Initialize the data vector x (or load data)


Let us plot the data and the true components before running the EM algorithm.

```{r}
#| label: lst-em-plot-synthtic-dataset
#| lst-label: lst-em-example-1
#| lst-cap: "Plotting the synthetic dataset"

# Plot the true density
par(mfrow=c(1,1))
xx.true = seq(-8,11,length=200)
yy.true = w.true*dnorm(xx.true, mu.true[1], sigma.true) + 
          (1-w.true)*dnorm(xx.true, mu.true[2], sigma.true) 
plot(xx.true, yy.true, type="l", xlab="x", ylab="True density", lwd=2)
points(x, rep(0,n), col=cc)
```

```{r}
#| label: lst-em-Location-Mixture-EM-plotting-helper
#| lst-label: lst-em-Location-Mixture-EM-plotting-helper
#| lst-cap: "Helper function for plotting the EM algorithm for a location mixture of Gaussians"

plot_components <- function(  mu, sigma, w, xx, yy.true, x, cc, s, QQ.out){
  # Plot current estimate over data
  layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
  par(mar=c(3.1,4.1,0.5,0.5))
  plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q", lwd=2)
  
  par(mar=c(5,4,1.5,0.5))
  xx = seq(-8,11,length=200)
  yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
  plot(xx, yy, type="l", ylim=c(0, max(c(yy,yy.true))), main=paste("s =",s,"   Q =", round(QQ.out[s],4)), lwd=2, col="red", lty=2, xlab="x", ylab="Density")
  lines(xx.true, yy.true, lwd=2)
  points(x, rep(0,n), col=cc)
  legend(6,0.22,c("Truth","Estimate"),col=c("black","red"), lty=c(1,2))
}
```

Now it is time to run the actual EM algorithm - in @lst-em-Location-Mixture-EM-Initialization we initialize the parameters of the algorithm.

```{r}
#| label: lst-em-Location-Mixture-EM-Initialization
#| lst-label: lst-em-Location-Mixture-EM-Initialization
#| lst-cap: "Initialization of the EM algorithm for a location mixture of Gaussians"

## Initialize the parameters
w     = 1/2                         # Assign equal weight to each component to start with
mu    = rnorm(KK, mean(x), sd(x))   # Random cluster centers randomly spread over the support of the data
sigma = sd(x)                       # Initial standard deviation

s  = 0
sw = FALSE
QQ = -Inf
QQ.out = NULL
epsilon = 10^(-5)
```

Mow we can Plot the initial guess for the density

```{r}
#| label: lst-em-Location-Mixture-EM-Initialization-plot
#| lst-label: lst-em-Location-Mixture-EM-Initialization-plot
#| lst-cap: "Initial parameter estimates of the EM algorithm for a location mixture of Gaussians"

xx = seq(-8,11,length=200)
yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
plot(xx, yy, type="l", ylim=c(0, max(yy)), xlab="x", ylab="Initial density")
points(x, rep(0,n), col=cc)
```


in @lst-em-Location-Mixture-EM we run the E and M steps until convergence.

```{r}
#| label: lst-em-Location-Mixture-EM
#| lst-label: lst-em-Location-Mixture-EM
#| lst-cap: "EM algorithm for a location mixture of Gaussians"

## Checking convergence of the algorithm
while(!sw){
  ## E step
  v = array(0, dim=c(n,KK))
  v[,1] = log(w) + dnorm(x, mu[1], sigma, log=TRUE)    # Compute the log of the weights
  v[,2] = log(1-w) + dnorm(x, mu[2], sigma, log=TRUE)  # Compute the log of the weights
  for(i in 1:n){
    v[i,] = exp(v[i,] - max(v[i,]))/sum(exp(v[i,] - max(v[i,])))  #Go from logs to actual weights in a numerically stable manner
  }
  
  ## M step
  # Weights
  w = mean(v[,1])
  mu = rep(0, KK)
  for(k in 1:KK){
    for(i in 1:n){
      mu[k]    = mu[k] + v[i,k]*x[i]
    }
    mu[k] = mu[k]/sum(v[,k])
  }
  # Standard deviations
  sigma = 0
  for(i in 1:n){
    for(k in 1:KK){
      sigma = sigma + v[i,k]*(x[i] - mu[k])^2
    }
  }
  sigma = sqrt(sigma/sum(v))
  
  ##Check convergence
  QQn = 0 # This is the value of the Q function at the new parameter estimates
  for(i in 1:n){
    QQn = QQn + v[i,1]*(log(w)   + dnorm(x[i], mu[1], sigma, log=TRUE)) +
                v[i,2]*(log(1-w) + dnorm(x[i], mu[2], sigma, log=TRUE))
  }
  if(abs(QQn-QQ)/abs(QQn)<epsilon){
    sw=TRUE
  }
  QQ = QQn
  QQ.out = c(QQ.out, QQ)
  s = s + 1
  # print(paste(s, QQn))
  plot_components(mu, sigma, w, xx, yy.true, x, cc, s, QQ.out)

}
```

```{r}
#| label: lst-em-final-estimate
#| lst-label: lst-em-final-estimate
#| lst-cap: "Final estimate of the EM algorithm"

# Plot final estimate over data
layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
par(mar=c(3.1,4.1,0.5,0.5))
plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q", lwd=2)

par(mar=c(5,4,1.5,0.5))
xx = seq(-8,11,length=200)
yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
plot(xx, yy, type="l", ylim=c(0, max(c(yy,yy.true))), main=paste("s =",s,"   Q =", round(QQ.out[s],4)), lwd=2, col="red", lty=2, xlab="x", ylab="Density")
lines(xx.true, yy.true, lwd=2)
points(x, rep(0,n), col=cc)
legend(6,0.22,c("Truth","Estimate"),col=c("black","red"), lty=c(1,2), bty="n")
```